---
title: "Human Disease Blood Atlas Olink Data Quality Control"
subtitle: "Phase 2 (Olink HT) - Batch 2"
author: "María Bueno Álvez"
date: last-modified
date-format: "YYYY-MM-DD"
format:
  html:
    self-contained: true
    theme: flatly
    title-block-banner: true
    smooth-scroll: true
    toc: true
    toc-depth: 4
    toc-location: right
    number-sections: true
    number-depth: 4
    code-fold: true
    code-tools: true
    code-copy: true
    code-overflow: wrap
    df-print: kable
    standalone: false
    fig-align: left
    figure:
      caption: "Figure {number}: {label}"
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
# Read internal functions
source("../../scripts/functions/functions_utility.R")
source("../../scripts/functions/functions_olink_qc.R")
source("../../scripts/functions/functions_visualization.R")
source("../../scripts/functions/themes_palettes.R")

# Load Olink NPX data and LIMS manifest
data_ht <-read_NPX("../../data/VL-3530-B2/VL-3530_B2_NPX_2025-05-08.parquet")
manifest <- import_df("../../data/samples_2025-05-12.xlsx")
```


```{r, include=FALSE}
# Extract QC information from Olink file
qc <- generate_qc_params(data = data_ht,
                         manifest = manifest)

# Extract dataset from QC
data_ht_lod <- qc$lod_data
  
## Save raw data - including LOD calculation
#rite_tsv(data_ht_lod, file = "../../data/final_data/data_phase2_batch2_raw_20250702.tsv")
```


# Data overview

This section provides an **overview of the Olink dataset**, including the total number of samples and proteins measured. 

In the delivered Olink dataset, there are currently **`r qc$n_samples` samples**, and **`r nrow(qc$olink_meta)` proteins** measured in each sample.

**Figure 1**: Number of samples per cohort.

```{r}
plot_cohort_n(plot_data = qc$manifest_filtered,
              save_plot = "da_phase2_batch2_cohorts")

plot_disease_n(plot_data = qc$manifest_filtered,
               save_plot = "da_phase2_batch2_cohorts")
```

**Figure 2**: Sample distribution across plates.

```{r}
plot_plate_distribution(plot_data = qc$mapped_data,
                        save_plot = "da_phase2_batch2_plates")
```


# Step 1: Exploration of control samples and control assays

This section identifies and **analyzes control samples to ensure data quality**. Specifically, we explore correlations between technical replicates for selected patients (patient 2 and patient 3) across multiple plates.


## Check missingness

```{r}
# In general
plot_NPX_missingness(plot_data = data_ht,
                     plot_title = "All samples")

# Control samples
plot_NPX_missingness(plot_data = data_ht,
                     samples = qc$control_samples,
                     plot_title = "Control samples")
```



## Plate-to-plate correlation for patient 2 & 3

We calculate **Spearman correlations for measurements of control samples (Patient 1, Patient 2, Patient 3 and Patient 5 across plates**. These analyses help assess technical reproducibility.

**Figure 3**: Heatmaps of correlations for control patients across plates (all measured proteins).

```{r}
plot_patient_heatmap(plot_data = data_ht,
                     select_id = qc$control_samples)
```


### Correlation excluding proteins below LOD

This section refines the **correlation analysis by excluding proteins that fall below the LOD across all replicates**. This step ensures a cleaner analysis focusing only on reliably measured proteins.


**Figure 4**: Heatmaps of correlations for control patients across plates (filtered for proteins above LOD).

```{r}
data_excluded <- 
  data_ht |> 
  filter(!Assay %in% qc$exclude_proteins$Assay) 

plot_patient_heatmap(plot_data = data_excluded,
                     select_id = qc$control_samples)
```


## Remove control assays & samples

In this step, we **filter out control assays and samples** that are not part of the main dataset. This ensures clean data for downstream analyses.

```{r}
data_step_1 <- 
  data_ht_lod |> 
  filter(!SampleType %in% qc$control_sample_types,
         !DAid %in% qc$control_samples, # Remove internal KTH controls
         !AssayType %in% qc$control_assay_types) |>
  select(DAid, Assay, OlinkID, UniProt, Panel, Block, Plate, Count, ExtNPX, NPX, PCNormalizedNPX, AssayQC, SampleQC, LOD) |> 
  mutate(above_LOD = ifelse(NPX > LOD, "Yes", "No"))
```


# Step 2: QC warnings

In this step, we assess **quality control (QC) warnings at the sample and protein levels**. Samples or protein measurements flagged with warnings will be excluded based on predefined thresholds.

## Percentage of QC warnings per sample
We calculate the number of QC warnings per sample, with a focus on those that do not pass QC.

**Figure 5**: Barplot showing the number of QC warnings for each sample, colored by cohort (top) and type of warning (bottom).

```{r}
plot_qc_warnings(plot_data = data_step_1,
                 plot_type = "barplot_cohort",
                 save_plot = "samples_qc_batch2_cohort")

plot_qc_warnings(plot_data = data_step_1,
                 plot_type = "barplot_type",
                 save_plot = "samples_qc_batch2_warning")
```

**Figure 6**: Beeswarm plot showing the number of proteins passing QC for each sample.

```{r}
plot_qc_warnings(plot_data = data_step_1,
                 plot_type = "beeswarm",
                 save_plot = "samples_qc_batch2_beeswarm")
```


## Samples with > 50% QC warnings

This analysis identifies samples with more than 50% of proteins flagged with QC warnings, which will be excluded from further analysis.

```{r}
samples_high_warnings <- calculate_sample_warnings(data = data_step_1)
```

In the current dataset, the **number of samples with > 50% warnings is: `r nrow(samples_high_warnings)`**.

These samples will be excluded from further analysis.

## Remove individual sample QC warnings

In this step, we **filter out the samples identified in the previous step that have more than 50% QC warnings**. Additionally, we **remove protein measurements that failed QC**. Since the data is in long format, this step will not introduce any NAs. However, when the data is reshaped into a wide format, missing values may be introduced.

```{r}
data_step_2 <- 
  data_step_1 |> 
  filter(SampleQC == "PASS",
         !DAid %in% samples_high_warnings$DAid) |>
  select(-SampleQC)
```

# Step 3: Assay warnings

In this step, we evaluate **quality control (QC) warnings at the assay level**. Assays with a high percentage of warnings will be flagged and excluded from the dataset based on a predefined threshold to ensure data reliability.

**Figure 7**: Bar plot showing the number of assays with warnings across different assays.

```{r}
plot_assay_warnings(plot_data = data_step_2, 
                    save_plot= "assay_qc_batch2")

assays_high_warnings <- calculate_assay_warnings(data = data_step_2)
```

In the current dataset, the number of **assays with > 50% warnings is: `r nrow(assays_high_warnings)`**.  

These assays will be excluded from further analysis.

```{r}
data_step_3 <- 
  data_step_2 |> 
  filter(!Assay %in% assays_high_warnings$Assay) |>
  select(-AssayQC)
```

# Step 4: Protein detectability

In this step, we calculate the **number of proteins detected above LOD per sample**, in order to explore potential differences in detectability across cohorts.

```{r}
sample_detected_proteins <- calculate_sample_detectability(data = data_step_3)
```

**Figure 8**: Beeswarm plot showing the number of proteins above LOD for each sample.

```{r}
plot_sample_detectability(plot_data = sample_detected_proteins,
                          save_plot = "detectability_batch2")
```

# Step 5: Outlier exploration

In this step, we use **PCA and UAMP to explore potential outliers in the dataset**. By visualizing the data in lower dimensions, we can identify samples that may behave differently from the rest and may need further investigation or exclusion.


```{r}
# Prepare data
input_data <- 
  data_step_3 |>
  select(DAid, OlinkID, NPX) |>
  pivot_wider(names_from = "OlinkID", values_from = "NPX")

input_data_filt <- 
  data_step_3 |>
  filter(!Assay %in% qc$exclude_proteins$Assay) |> 
  select(DAid, OlinkID, NPX) |>
  pivot_wider(names_from = "OlinkID", values_from = "NPX")
```


## PCA


**Figure 9**: PCA visualization based on all proteins, where samples are colored by 1) class, 2) Disease, 3) plate, 4) number of detected proteins.

```{r}
pca_all_proteins <-
  do_pca(data = input_data,
          plots = F)

plot_dim_reduction(plot_data = pca_all_proteins$pca_res,
                   plot_type = "class",
                   plot_title = "PCA all proteins (colored by Class)",
                   method = "pca",
                   save_plot = "p2_b2_pca_all_class") 

plot_dim_reduction(plot_data = pca_all_proteins$pca_res,
                   plot_type = "disease",
                   plot_title = "PCA all proteins (colored by Disease)",
                   method = "pca",
                   save_plot = "p2_b2_pca_all_disease") 

plot_dim_reduction(plot_data = pca_all_proteins$pca_res,
                   plot_type = "plate",
                   plot_title = "PCA all proteins (colored by Plate)",
                   method = "pca",
                   save_plot = "p2_b2_pca_all_plate") 

plot_dim_reduction(plot_data = pca_all_proteins$pca_res,
                   plot_type = "detectability",
                   plot_title = "PCA all proteins (colored by Detectability)",
                   method = "pca",
                   save_plot = "p2_b2_pca_all_detectability") 
```

**Figure 10**: PCA visualization based on proteins with high detectability, where samples are colored by disease

```{r}
pca_high_detectability <-
  do_pca(data = input_data_filt,
         plots = F)

plot_dim_reduction(plot_data = pca_high_detectability$pca_res,
                   plot_type = "disease",
                   plot_title = "PCA high detectability (colored by Disease)",
                   method = "pca",
                   save_plot = "p2_b2_pca_high_detectability_disease") 
```


## UMAP

**Figure 11**: UMAP visualization based on all proteins, where samples are colored by 1) class, 2) Disease, 3) plate, 4) number of detected proteins.

```{r}
umap_all_proteins <-
  do_umap(data = input_data,
          plots = F)

plot_dim_reduction(plot_data = umap_all_proteins,
                   plot_type = "class",
                   plot_title = "UMAP all proteins (colored by Class)",
                   method = "umap",
                   save_plot = "p2_b2_umap_all_class") 

plot_dim_reduction(plot_data = umap_all_proteins,
                   plot_type = "disease",
                   plot_title = "UMAP all proteins (colored by Disease)",
                   method = "umap",
                   save_plot = "p2_b2_umap_all_disease") 

plot_dim_reduction(plot_data = umap_all_proteins,
                   plot_type = "plate",
                   plot_title = "UMAP all proteins (colored by Plate)",
                   method = "umap",
                   save_plot = "p2_b2_umap_all_plate") 

plot_dim_reduction(plot_data = umap_all_proteins,
                   plot_type = "detectability",
                   plot_title = "UMAP all proteins (colored by Detectability)",
                   method = "umap",
                   save_plot = "p2_b2_umap_all_detectability") 
```


**Figure 12**: UMAP visualization based on proteins with high detectability, where samples are colored by disease

```{r}
umap_high_detectability <-
  do_umap(data = input_data_filt,
          plots = F)

plot_dim_reduction(plot_data = umap_high_detectability,
                   plot_type = "disease",
                   plot_title = "UMAP high detectability (colored by Disease)",
                   method = "umap",
                   save_plot = "p2_b2_umap_high_detectability_disease") 
```



# Save final data

After completing the quality control and data curation steps, we **save the cleaned and processed dataset for sharing with collaborators and for further downstream analysis**. This final dataset is ready for use in subsequent phases of the project.

```{r}
curated_data <- 
  data_step_3 |> 
  select(-Count,	-ExtNPX, -PCNormalizedNPX) # Remove unnecessary columns

final_n <- calculate_n_samples_assays(curated_data)

write_tsv(curated_data, file = "../../data/final_data/data_phase2_batch2_curated_20250702.tsv") 
```

In the final dataset, there are  **`r final_n$n_samples` samples**, and **`r final_n$n_proteins` proteins**.
