---
title: "Human Disease Blood Atlas Olink Data Quality Control"
subtitle: "Phase 2 (Olink HT) - Batch 5"
author: "María Bueno Álvez"
date: last-modified
date-format: "YYYY-MM-DD"
format:
  html:
    self-contained: true
    theme: flatly
    title-block-banner: true
    smooth-scroll: true
    toc: true
    toc-depth: 4
    toc-location: right
    number-sections: true
    number-depth: 4
    code-fold: true
    code-tools: true
    code-copy: true
    code-overflow: wrap
    df-print: kable
    standalone: false
    fig-align: left
    figure:
      caption: "Figure {number}: {label}"
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
# Load packages
library(tidyverse) 
library(ggplotify)
library(patchwork)
library(pheatmap)
library(ggbeeswarm)
library(ggrepel)
library(GGally)
library(arrow)
library(OlinkAnalyze)

# Read internal functions
source("../../scripts/functions/functions_utility.R")
source("../../scripts/functions/functions_visualization.R")
source("../../scripts/functions/themes_palettes.R")

list.files("../data")

# Load Olink NPX data and LIMS manifest
data_ht <- read_NPX("../../data/VL-3530-B5/VL-3530B5_NPX_2025-05-08.parquet")
manifest <- import_df("../../data/samples_2025-05-12.xlsx")
```


```{r, include=FALSE}
# Define control sample type, control assay types, replicate samples, replicate assays
control_sample_types <- c("SAMPLE_CONTROL", "PLATE_CONTROL", "NEGATIVE_CONTROL")
control_assay_types <- c("ext_ctrl", "inc_ctrl", "amp_ctrl")

n_samples_run <- 
  data_ht |> 
  filter(!SampleType %in% control_sample_types) |>
  distinct(SampleID) |> 
  nrow()

control_samples <- 
  manifest |> 
  filter(Cohort == "CTRL") |> 
  pull(DAid)

# Extract Olink assay metadata 
olink_meta <- 
  data_ht |> 
  filter(!AssayType %in% control_assay_types) |>
  distinct(Assay, OlinkID, UniProt, AssayType, Block)

n_proteins <- 
  olink_meta$Assay |>
  unique() |> 
  length()

# Map SampleIDs to DA identifiers 
id_mapping <- 
  data_ht |> 
  filter(SampleType == "SAMPLE") |> 
  distinct(SampleID) |> 
  separate(SampleID, into = c("DAid", "Batch", "Plate"), sep = "-", remove = F) 

data_hdba <- 
  data_ht |>
  left_join(id_mapping, by = "SampleID") |>
  relocate(DAid)

n_samples <- 
  data_hdba |> 
  filter(!SampleType %in% control_sample_types) |> 
  distinct(DAid) |> 
  nrow()

# Calculate Limit of Detection (LOD) using negative controls
data_ht_lod <- olink_lod(data = data_hdba, lod_method = "NCLOD")

## Save raw data - including LOD calculation
getwd()
write_tsv(data_ht_lod, file = "../../data/final_data/data_phase2_batch5_raw_20250624.tsv")
```


# Data overview

This section provides an **overview of the Olink dataset**, including the total number of samples and proteins measured. 

In the delivered Olink dataset, there are currently **`r n_samples` samples**, and **`r nrow(olink_meta)` proteins** measured in each sample.

**Figure 1**: Number of samples per cohort.

```{r}
data_hdba |> 
  filter(!SampleType %in% control_sample_types) |> 
  distinct(DAid) |> 
  left_join(manifest, by = "DAid") |> 
  mutate(Cohort = paste(Cohort, `PI - sample owner`, sep = "_")) |>
  count(Cohort) |> 
  ggplot(aes(Cohort, n, fill = Cohort)) +
  geom_col() +
    scale_fill_manual(values = brewer.pal(n = 11, name = "Set3")) +  
  geom_text(aes(label = n), vjust = -0.5) +
  theme_hpa(angled = T) +
  ylab("Number of samples")

ggsave(savepath("da_phase2_batch5_cohorts.png"), h = 6, w = 6)
```

**Figure 2**: Sample distribution across plates.

```{r}
data_hdba |> 
  mutate(Plate = gsub("P", "", Plate),
         Plate = as.numeric(Plate),
         Plate = as.factor(Plate)) |>
  distinct(DAid, SampleID, Plate, SampleType) |> 
  filter(!is.na(Plate)) |> 
  left_join(manifest, by = "DAid") |> 
  mutate(Cohort = paste(Cohort, `PI - sample owner`, sep = "_"),
         Cohort = ifelse(SampleType %in% control_sample_types, "CONTROLS", Cohort))  |> 
  group_by(Plate, Cohort) |> 
  count() |> 
  ggplot(aes(Plate, n, fill = Cohort)) +
  geom_col() +
  scale_fill_manual(values = brewer.pal(n = 11, name = "Set3")) +  
  theme_hpa(angled = T) +
  ylab("Number of samples")

ggsave(savepath("da_phase2_batch5_plates.png"), h = 6, w = 8)
```


# Step 1: Exploration of control samples and control assays

This section identifies and **analyzes control samples to ensure data quality**. Specifically, we explore correlations between technical replicates for selected patients (patient 2 and patient 3) across multiple plates.

## Plate-to-plate correlation for patient 2 & 3

We calculate **Spearman correlations for measurements of Patient 2 and Patient 3 across plates**. These analyses help assess technical reproducibility.



**Figure 3**: Correlation heatmap for Patient 2 across plates.

```{r}
data_ht |> 
  left_join(id_mapping, by = "SampleID") |> 
  filter(DAid == "DA12362") |> 
  select(Plate, OlinkID, PCNormalizedNPX) |> 
  pivot_wider(names_from = OlinkID, values_from = PCNormalizedNPX) |> 
  column_to_rownames("Plate") |> 
  t() |> 
  cor(method = "spearman")|> 
  pheatmap()|> 
  as.ggplot()

#ggsave(savepath("p2_heatmap.png"), h = 8, w = 8)
```

**Figure 4**: Correlation heatmap for Patient 3 across plates.

```{r}
data_ht |>  
  left_join(id_mapping, by = "SampleID") |> 
  filter(DAid == "DA12363") |> 
  select(Plate, OlinkID, NPX) |> 
  pivot_wider(names_from = OlinkID, values_from = NPX) |> 
  column_to_rownames("Plate") |> 
  t() |> 
  cor(method = "spearman")|> 
  pheatmap() |> 
  as.ggplot()

#ggsave(savepath("p3_heatmap.png"), h = 8, w = 8)
```


**Figure 5**: Scatterplots showing pairwise correlations between plates for Patient 2.

```{r warning=FALSE, message=FALSE}
data_ht |> 
  left_join(id_mapping, by = "SampleID") |> 
  filter(DAid == "DA12362") |> 
  select(Plate, OlinkID, NPX) |> 
  pivot_wider(names_from = Plate, values_from = NPX) |> 
  ggpairs(c(2:11),
            lower = list(continuous = "points", size = 0.05)) +
  ggtitle("Correlation for Patient 2 measurements across plates") 

#ggsave(savepath("p2_plate_scatterplots.png"), h = 10, w = 10)
```

**Figure 6**: Scatterplots showing pairwise correlations between plates for Patient 3.

```{r warning=FALSE, message=FALSE}
data_ht |> 
  left_join(id_mapping, by = "SampleID") |> 
  filter(DAid == "DA12363") |> 
  select(Plate, OlinkID, NPX) |> 
  pivot_wider(names_from = Plate, values_from = NPX) |> 
  ggpairs(c(2:11),
            lower = list(continuous = "points", size = 0.5)) +
  ggtitle("Correlation for Patient 3 measurements across plates")

#ggsave(savepath("p3_plate_scatterplots.png"), h = 10, w = 10)
```

### Correlation excluding proteins below LOD

This section refines the **correlation analysis by excluding proteins that fall below the LOD across all replicates**. This step ensures a cleaner analysis focusing only on reliably measured proteins.

```{r warning=FALSE, message=FALSE}
exclude_proteins <- 
  data_ht_lod |> 
  filter(DAid %in% control_samples) |> 
  mutate(under_LOD = ifelse(NPX < LOD, "Yes", "No")) |> 
  group_by(Assay) |> 
  count(under_LOD) |> 
  filter(under_LOD == "Yes") |> 
  ungroup() |> 
  filter(n > length(unique(data_ht$PlateID)))
```

**Figure 7**: Heatmaps of correlations for Patient 2 measurements (filtered for proteins above LOD).

```{r}
data_ht |> 
  left_join(id_mapping, by = "SampleID") |> 
  filter(DAid == "DA12362",
         !Assay %in% exclude_proteins$Assay) |> 
  select(Plate, OlinkID, PCNormalizedNPX) |> 
  pivot_wider(names_from = OlinkID, values_from = PCNormalizedNPX) |> 
  column_to_rownames("Plate") |> 
  t() |> 
  cor(method = "spearman")|> 
  pheatmap() |> 
  as.ggplot()

#ggsave(savepath("p2_subset_heatmap.png"), h = 8, w = 8)
```

**Figure 8**: Heatmaps of correlations for Patient 3 measurements (filtered for proteins above LOD).

```{r}
data_ht |> 
  left_join(id_mapping, by = "SampleID") |> 
  filter(DAid == "DA12363",
         !Assay %in% exclude_proteins$Assay) |> 
  select(Plate, OlinkID, NPX) |> 
  pivot_wider(names_from = OlinkID, values_from = NPX) |> 
  column_to_rownames("Plate") |> 
  t() |> 
  cor(method = "spearman")|> 
  pheatmap() |> 
  as.ggplot()

#ggsave(savepath("p3_subset_heatmap.png"), h = 8, w = 8)
```


## Remove control assays & samples

In this step, we **filter out control assays and samples** that are not part of the main dataset. This ensures clean data for downstream analyses.

```{r}
data_step_1 <- 
  data_ht_lod |> 
  filter(!SampleType %in% control_sample_types,
         !DAid %in% control_samples, # Remove internal KTH controls
         !AssayType %in% control_assay_types) |>
  select(DAid, Assay, OlinkID, UniProt, Panel, Block, Plate, Count, ExtNPX, NPX, PCNormalizedNPX, AssayQC, SampleQC, LOD) |> 
  mutate(above_LOD = ifelse(NPX > LOD, "Yes", "No"))
```


# Step 2: QC warnings

In this step, we assess **quality control (QC) warnings at the sample and protein levels**. Samples or protein measurements flagged with warnings will be excluded based on predefined thresholds.

## Percentage of QC warnings per sample
We calculate the number of QC warnings per sample, with a focus on those that do not pass QC.

**Figure 9**: Barplot showing the number of QC warnings for each sample, colored by cohort.

```{r}
data_step_1 |> 
  group_by(DAid) |> 
  count(SampleQC) |> 
  filter(!SampleQC %in% c("NA", "PASS")) |> 
  arrange(-n) |> 
  left_join(manifest, by = "DAid") |> 
  mutate(Cohort = paste(Cohort, `PI - sample owner`, sep = "_")) |>
  ggplot(aes(fct_reorder(DAid, -n), n, fill = Cohort)) +
  geom_col() +
  scale_fill_manual(values = brewer.pal(n = 11, name = "Set3")) +  
  geom_hline(yintercept = 5440/2, linetype = "dashed", color = "grey") +
  theme_hpa(angled = T) +
  xlab("Sample") +
  theme(axis.text.x = element_text(size = 5),
        axis.ticks.x = element_blank()) +
  ggtitle("Samples with QC warnings") 


ggsave(savepath("samples_qc_batch5.png"), h = 4, w = 10)

data_step_1 |> 
  group_by(DAid) |> 
  count(SampleQC) |> 
  filter(!SampleQC %in% c("NA", "PASS")) |> 
  arrange(-n) |> 
  left_join(manifest, by = "DAid") |> 
  mutate(Cohort = paste(Cohort, `PI - sample owner`, sep = "_")) |>
  ggplot(aes(fct_reorder(DAid, -n), n, fill = SampleQC)) +
  geom_col() +
  geom_hline(yintercept = 5440/2, linetype = "dashed", color = "grey") +
  theme_hpa(angled = T) +
  xlab("Sample") +
  theme(axis.text.x = element_text(size = 5),
        axis.ticks.x = element_blank()) +
  ggtitle("Samples with QC warnings") 

ggsave(savepath("samples_qc_2_batch5.png"), h = 4, w = 10)
```

**Figure 10**: Beeswarm plot showing the number of proteins passing QC for each sample.

```{r}
data_step_1 |> 
  group_by(DAid) |> 
  filter(SampleQC == "PASS") |>
  count(SampleQC) |> 
  left_join(manifest, by = "DAid") |> 
  filter(Cohort != "CTRL") |> 
  mutate(Cohort = paste(Cohort, `PI - sample owner`, sep = "_")) |>
  ggplot(aes(Cohort, n, color = Cohort)) +
  geom_quasirandom() +
  geom_hline(yintercept = n_proteins / 2, color = "grey40", linetype = "dashed") +
 # geom_text_repel(aes(label = DAid), show.legend = F) +
  scale_color_manual(values = pal_phase2) +
  theme_hpa(angled = T) +
  ggtitle("Number of proteins that pass QC per sample")

ggsave(savepath("samples_pass_batch5.png"), h = 8, w = 8)
```


## Samples with > 50% QC warnings

This analysis identifies samples with more than 50% of proteins flagged with QC warnings, which will be excluded from further analysis.

```{r}
samples_high_warnings <- 
  data_step_1 |> 
  filter(SampleQC != "PASS") |> 
  count(DAid) |> 
  mutate(perc_warning = n / n_proteins) |> 
  filter(perc_warning > 0.5) 
```

In the current dataset, the **number of samples with > 50% warnings is: `r nrow(samples_high_warnings)`**.

These samples will be excluded from further analysis.

## Remove individual sample QC warnings

In this step, we **filter out the samples identified in the previous step that have more than 50% QC warnings**. Additionally, we **remove protein measurements that failed QC**. Since the data is in long format, this step will not introduce any NAs. However, when the data is reshaped into a wide format, missing values may be introduced.

```{r}
data_step_2 <- 
  data_step_1 |> 
  filter(SampleQC == "PASS",
         !DAid %in% samples_high_warnings$DAid) |>
  select(-SampleQC)
```

# Step 3: Assay warnings

In this step, we evaluate **quality control (QC) warnings at the assay level**. Assays with a high percentage of warnings will be flagged and excluded from the dataset based on a predefined threshold to ensure data reliability.

**Figure 11**: Bar plot showing the number of assays with warnings across different assays.

```{r}
data_step_2 |>
  group_by(Assay) |> 
  count(AssayQC) |> 
  filter(!AssayQC %in% c("NA", "PASS")) |> 
  arrange(-n) |> 
  ggplot(aes(fct_reorder(Assay,-n), n)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.5) +
  theme_hpa(angle = T) +
  xlab("Assay")

ggsave(savepath("assay_qc_batch5.png"), h = 5, w = 10)

assays_high_warnings <- 
  data_step_2 |> 
  filter(!DAid %in% control_samples) |> 
  group_by(Assay) |> 
  mutate(n_total = n_distinct(DAid)) |> 
  filter(AssayQC == "WARN") |> 
  mutate(n = n_distinct(DAid)) |> 
  ungroup() |> 
  distinct(Assay, n, n_total) |> 
  mutate(perc_warning = n / n_total) |> arrange(-n) |> 
  filter(perc_warning > 0.5) 
```

In the current dataset, the number of **assays with > 50% warnings is: `r nrow(assays_high_warnings)`**.  

These assays will be excluded from further analysis.

```{r}
data_step_3 <- 
  data_step_2 |> 
  filter(!Assay %in% assays_high_warnings$Assay) |>
  select(-AssayQC)
```

# Protein detectability

```{r}
disease_order <- 
  manifest |> 
  mutate(Disease = ifelse(Disease == "Healthy", paste(Disease, Cohort, sep = " - "), Disease)) |> 
  filter(DAid %in% data_step_3$DAid) |> 
  distinct(Disease, Cohort, Class) |> 
  arrange(desc(Cohort))

sample_detected_proteins |> 
  left_join(manifest, by = "DAid") |> 
  mutate(Disease = ifelse(Disease == "Healthy", paste(Disease, Cohort, sep = " - "), Disease),
         Disease = factor(Disease, levels = disease_order$Disease)) |> 
  ggplot(aes(Disease, n, fill = Cohort, color = Cohort)) +
  geom_quasirandom() +
  geom_boxplot(outlier.color = NA, alpha = 0.5, color = "black") +
  scale_fill_manual(values = brewer.pal(n = 11, name = "Set3")) +  
  scale_color_manual(values = brewer.pal(n = 11, name = "Set3")) +  
  theme_hpa(angled = T)

ggsave(savepath("detectability_batch5.png"), h = 5, w = 10)
```

# Step 4: Outlier exploration

In this step, we use **PCA and UAMP to explore potential outliers in the dataset**. By visualizing the data in lower dimensions, we can identify samples that may behave differently from the rest and may need further investigation or exclusion.


```{r}
# Prepare data
input_data <- 
  data_step_3 |>
  select(DAid, OlinkID, NPX) |>
  pivot_wider(names_from = "OlinkID", values_from = "NPX")

# Exclude protens below LOD in 50% of control measurements
exclude_proteins <- 
  data_ht_lod |> 
  filter(DAid %in% control_samples) |> 
  mutate(under_LOD = ifelse(NPX < LOD, "Yes", "No")) |>
  group_by(Assay) |> 
  count(under_LOD) |> 
  filter(under_LOD == "Yes") |> 
  ungroup() |> 
  filter(n >= sum(!is.na(unique(data_ht_lod$Plate))))

input_data_filt <- 
  data_step_3 |>
  filter(!Assay %in% exclude_proteins$Assay) |> 
  select(DAid, OlinkID, NPX) |>
  pivot_wider(names_from = "OlinkID", values_from = "NPX")
```


## PCA


**Figure 15**: PCA visualization based on all proteins, where samples are colored by cohort.

```{r}
pca_all_proteins <-
  do_pca(data = input_data,
          plots = F)

# plot_pca(pca_results = pca_all_proteins,
#          color = "Class")
# 
# plot_pca(pca_results = pca_all_proteins,
#          color = "Disease")
# 
# plot_pca(pca_results = pca_all_proteins,
#          color = "Plate")
# 
# plot_pca(pca_results = pca_all_proteins,
#          color = "Detectability",
#          data = )


pca_all_proteins$pca_res |>
  rename(DAid = Sample) |>
  left_join(manifest, by = "DAid") |>
  mutate(Cohort = paste(Cohort, `PI - sample owner`, sep = "_")) |>
  ggplot(aes(PC1, PC2, color = Class)) +
  geom_point(alpha = 0.7, size = 0.8) +
  scale_color_manual(values = pal_class_2) +
  theme_hpa() 

ggsave(savepath("b5_pca_class.png"), h = 8, w = 8)

pca_all_proteins$pca_res |>
  rename(DAid = Sample) |>
  left_join(manifest, by = "DAid") |>
  ggplot(aes(PC1, PC2, color = Disease)) +
  geom_point(alpha = 0.7, size = 0.8) +
  theme_hpa() 

ggsave(savepath("b5_pca_disease.png"), h = 8, w = 8)

# Color by plate
pca_all_proteins$pca_res |>
  rename(DAid = Sample) |>
  left_join(manifest, by = "DAid") |>
  ggplot(aes(PC1, PC2, color = Disease)) +
  geom_point(alpha = 0.7, size = 0.8) +
  theme_hpa() 

ggsave(savepath("b5_pca_disease.png"), h = 8, w = 8)

# Color by detectability
pca_all_proteins$pca_res |>
  rename(DAid = Sample) |>
  left_join(manifest, by = "DAid") |>
  ggplot(aes(PC1, PC2, color = Disease)) +
  geom_point(alpha = 0.7, size = 0.8) +
  theme_hpa() 

ggsave(savepath("b5_pca_disease.png"), h = 8, w = 8)
```


**Figure 16**: PCA visualization based on proteins with high detectability, where samples are colored by cohort.

```{r}
pca_high_detectability <-
  do_pca(data = input_data_filt,
         plots = F)

pca_high_detectability$pca_res |>
  rename(DAid = Sample) |>
  left_join(manifest, by = "DAid") |>
  ggplot(aes(PC1, PC2, color = Disease)) +
  geom_point(alpha = 0.7, size = 0.8) +
  theme_hpa() +
  ggtitle("Proteins with high detectability")

ggsave(savepath("b5_pca_high_detectability.png"), h = 8, w = 8)
```


## UMAP

**Figure 17**: UMAP visualization based on all proteins, where samples are colored by cohort.

```{r}
umap_all_proteins <-
  do_umap(data = input_data,
          plots = F)

umap_all_proteins |>
  rename(DAid = Sample) |>
  left_join(manifest, by = "DAid") |>
  mutate(Cohort = paste(Cohort, `PI - sample owner`, sep = "_")) |>
  ggplot(aes(UMAP1, UMAP2, color = Class)) +
  geom_point(alpha = 0.7, size = 0.8) +
  scale_color_manual(values = pal_class_2) +
  theme_hpa() +
  ggtitle("All proteins")


umap_all_proteins |>
  rename(DAid = Sample) |>
  left_join(manifest, by = "DAid") |>
  left_join(sample_detected_proteins, by = "DAid") |> 
  mutate(Cohort = paste(Cohort, `PI - sample owner`, sep = "_")) |>
  ggplot(aes(UMAP1, UMAP2, color = Disease)) +
  geom_point(alpha = 0.7, size = 0.8) +
  #scale_color_manual(values = pal_class_2) +
 # scale_color_viridis_c() +
  theme_hpa() +
  ggtitle("All proteins")


```


**Figure 18**: UMAP visualization based on proteins with high detectability, where samples are colored by cohort.

```{r}
umap_high_detectability <-
  do_umap(data = input_data_filt,
          plots = F)

umap_high_detectability |>
  rename(DAid = Sample) |>
  left_join(manifest, by = "DAid") |>
#  mutate(Outlier = ifelse(DAid %in% umap_outliers$DAid, "Yes", "No")) |>  
  ggplot(aes(UMAP1, UMAP2, color = Disease)) +
  geom_point(alpha = 0.7, size = 0.8) +
 # scale_color_manual(values = c("Yes" = "red", "No" = "grey")) +
  theme_hpa() +
  ggtitle("Proteins with high detectability") 

ggsave(savepath("umap_high_detectability.png"), h = 8, w = 8)
```



# Save final data

After completing the quality control and data curation steps, we **save the cleaned and processed dataset for sharing with collaborators and for further downstream analysis**. This final dataset is ready for use in subsequent phases of the project.

```{r}
curated_data <- 
  data_step_3 |> 
  select(-Count,	-ExtNPX, -PCNormalizedNPX) # Remove unnecessary columns

write_tsv(curated_data, file = "../../data/final_data/data_phase2_batch5_curated_20250624.tsv") 
```

